---
title: "This Code is Big Time"
description: |
  Wrap your R code in Your Mom's House audio drops
date: "2021-04-25"
categories: [R, tutorial, Your Mom's House, programming]
---


What's up, Jeans. This short post is targeting the *very* niche crowd of R users who also happen to love Tim and Christine, the main mommies of Your Mom's House. Basically, I discovered the [`{beepr}` package](https://github.com/rasmusab/beepr) a few days ago, which makes it easier you play short sound files through R, and so naturally my first thought was to "try it out" with a YMH drop. The result was 2 functions added to my personal/miscellaneous package, [{`eemisc}`](https://github.com/ekholme/eemisc):

- `big_time()`, which will play the "THIS SHIT IS BIG TIME" drop from the [YMH intro](https://www.youtube.com/watch?v=r-6LamuOl8I), and
- `big_time_operator()`, which lets you wrap a function and play the "THIS SHIT IS BIG TIME" drop whenever you call that function.

I'll demonstrate these below.

# Setup

To use these functions, you'll need my personal package, `{eemisc}`, installed. You can install it via Github via the following command:

```{r install, eval = FALSE}
remotes::install_github("ekholme/eemisc")
```


Eventually, I might put these functions into their own package, but for now they live in my odds-and-ends package.

# Big Time

To play the "THIS SHIT IS BIG TIME" drop, you just need to call the function `big_time()`:

```{r setup}
library(eemisc)

big_time()
```


That's it. That will play the drop.

# Big Time Operator

But I took it a step further for all of my kings and queens above 18. The `big_time_operator()` takes a function of your choosing and produces as its output a new function that wraps the input function with the "big time" drop. Want to let everyone know that taking the mean is big time?

```{r big-time-mean}
big_time_mean <- big_time_operator(mean)

x <- 1:10

big_time_mean(x)
```


You can pass any function you want into `big_time_operator()` to get the same effect.

Anyway, that's all for now. Keep it high and tight, Jeans, and you bet I'm coming up in May.