---
title: "Crossfit Open '24 Analysis"
description: |
    An exploration of data from the 2024 Crossfit Open
date: 2024-03-19
execute:
    warning: false
format:
    html:
        code-fold: true
        fig-height: 8
        fig-width: 8
categories: [R, Crossfit, EDA]
---

Now that the 2024 Open is over, I figured it might be fun to look at some of the data from this year's top 100 finishers. There's a lot we can look at here, and I'm sure I'm going to leave a few things out -- feel free to drop me a line if there's something, like, urgent that I missed.

If you're into [R](https://www.r-project.org/) and want to do your own analysis, you can check out my work-in-progress [crossfitgames package](https://github.com/ekholme/crossfitgames) that has some tools for fetching and processing data from the CrossFit API. It's in kind of a janky state right now, but it works fine for what I want to do today.

I'm including the code used to pull/clean data and create graphs here. It'll be folded up by default, and I won't really explain what it's doing step-by-step, but if you're into that sort of thing, you can take a look.

```{r setup}
library(crossfitgames)
library(tidyverse)
library(gt)
library(hrbrthemes)
library(lubridate)

#placeholder for now
theme_set(theme_ipsum_rc())

women_raw <- open_leaderboard(2024, "women", top_n = 100)
men_raw <- open_leaderboard(2024, "men", top_n = 100)

women_lb <- extract_final_leaderboard(women_raw) |>
  mutate(div = "women")

men_lb <- extract_final_leaderboard(men_raw) |>
  mutate(div = "men")

#combine the two dataframes
all_lb <- rbind(women_lb, men_lb)

women_workout_res <- extract_workout_results(women_raw) |>
  mutate(div = "women")

men_workout_res <- extract_workout_results(men_raw) |>
  mutate(div = "men")

all_workouts <- rbind(women_workout_res, men_workout_res)
 
```


## Top Finishers

First, let's look at the top finishers for the men and women. We'll start with the women:

```{r}
women_lb |>
  slice_min(score, n = 10) |>
  select(-div) |> 
  gt() |> 
  tab_header(
    title = "Top 10 Women",
    subtitle = "2024 CF Open"
  )
```

One thing to keep in mind here is that lower scores are better in the Open, since your score is the sum of your rankings for each workout. Since there are 3 workouts, the best score possible would be 3.

On the women's side, the most recognizable names are Loewen and Saunders, and a lot of the bigger names (Tia, Laura, Emma Lawson) are absent. This shouldn't be concerning, but it's potentially interesting.

And moving to the men's side:

```{r}
men_lb |>
  slice_min(score, n = 10) |>
  select(-div) |> 
  gt() |> 
  tab_header(
    title = "Top 10 Men",
    subtitle = "2024 CF Open"
  )
```

The story is a little different here -- lots of noticeable names that have historically done well at the games. Vellner, Adler, Ohlsen, Koski, Panchik, and Crouch.

## Men and Women Scores by Rank 

Another way we might look at the data is to compare the scores of the men to those of the women. This can give us a sense of the "depth" or level of competition across the field.

For instance, we can plot each athlete's rank (overall placement) against the total score for men and women:

```{r}
ggplot(all_lb, aes(y = rank, x = score, color = div)) +
  geom_point(alpha = .5) +
  scale_y_reverse() +
  labs(
    x = "Score",
    y = "Rank (Overall Place)",
    title = "Score by Overall Rank",
    subtitle = "Top 100 Men and Women"
  ) +
  scale_color_discrete(name = "Division")
```

In cases where, for any given rank, the women's score is less than the men's score, we can assume (sort of) the women outperformed the men. For example, the 100th place woman had considerably fewer points than the 100th place man. This might tell us that, relative to the rest of the field, the 100th place woman is "better" than the 100th place man. Of course it's not that straightforward -- we could just as reasonably conclude that the men's field is deeper than the women's field -- but it's a fun thought exercise.

What is interesting, though, is that the score gap between the 1st and 100th woman is much smaller than the gap between the 1st and 100th point man. And that we see separation between the men and women around ~50th place.

## Variance in Workouts -- Women

Another interesting datapoint we can look at is the spread between an athlete's best finish and their worst finish. Since the open is only 3 events, this rank will explain a lot of their overall variance (at the Games, where there are more events, the best-to-worst spread is kinda less important). Obviously, all of these athletes finished top 100 overall, so nobody completely bombed anything.

Let's start with the women:

```{r}
women_best_worst <- women_workout_res |>
  group_by(athlete) |>
  summarize(
    worst = max(workout_place),
    best = min(workout_place),
    spread = worst - best
  ) |>
  left_join(women_lb, by = "athlete")

women_best_worst |>
  ggplot(aes(y = rank, x = best)) +
  geom_segment(aes(xend = worst, yend = rank), color = "steelblue") +
  scale_y_reverse() +
  labs(
    x = "Finish Position (Spread)",
    y = "Rank (Overall Place)",
    title = "Best and Worst Finishes by Overall Rank",
    subtitle = "Women"
  )
 
```

Ok, so the way you interpret this plot:

- The athlete's overall rank is on the y-axis (the vertical axis)
- The blue bar represents the spread of their performance. The left end is their best performance, the right end is their worst performance, and the width is the gap between best and worst

Keep in mind that having a small spread between your best and worst finish isn't *necessarily* a good thing if your best finish is relatively high.

All in all, though, we tend to see bigger spreads the further down we get in the rankings, which kinda makes sense because a single bad event can crush your overall score with just 3 total events.

From here, we might want to look at which women had the largest best-to-worst even spreads:


```{r}
women_best_worst |>
  slice_max(spread, n = 10) |>
  select(-div) |> 
  gt() |> 
  tab_header(
    title = "Largest Differences between Best and Worst Event",
    subtitle = "Women, CF Open 2024"
  )
```

We see Laura on here, and I think she did (relatively) poorly on the first event compared to her other 2 event finishes, hence the large spread:

```{r}
women_workout_res |> 
  filter(athlete == "Laura Horvath") |> 
  select(-div) |> 
  gt() |> 
  tab_header(
    title = "Laura Horvath Open Finishes"
  )
```

Right, so she finished 395th in event 1, but then 6th and 7th.

## Variance in Workouts -- Men 

And we can do the same thing for the men:

```{r}
men_best_worst <- men_workout_res |>
  group_by(athlete) |>
  summarize(
    worst = max(workout_place),
    best = min(workout_place),
    spread = worst - best
  ) |>
  left_join(men_lb, by = "athlete")

men_best_worst |>
  ggplot(aes(y = rank, x = best)) +
  geom_segment(aes(xend = worst, yend = rank), color = "steelblue") +
  scale_y_reverse() +
  labs(
    x = "Finish Position (Spread)",
    y = "Rank (Overall Place)",
    title = "Best and Worst Finishes by Overall Rank",
    subtitle = "Men"
  )
```


So the point about having a small spread not necessarily being ideal is illustrated here -- we see whoever finished in 98th has a teeny spread, but all of his finishes were ~250ish. 

```{r}
men_best_worst |>
  slice_max(spread, n = 10) |>
  gt() |> 
  tab_header(
    title = "Largest Differences between Best and Worst Event",
    subtitle = "Men, CF Open 2024"
  )

```

These spreads are wider than the ones we saw for the women, which again maybe suggests that the men's field is more variable than the women's field?

## 24.1

Now let's look briefly at individual workout results. Obviously, we'll start with 24.1. You can see the workout [here](https://games.crossfit.com/workouts/open/2024/1).


```{r}
wk1 <- all_workouts |>
  filter(workout_num == 1) |>
  mutate(time_score = period_to_seconds(ms(score)))

wk1 |>
  ggplot(aes(x = time_score, y = workout_place, color = div)) +
  geom_point(alpha = .5) +
  scale_y_reverse() +
  scale_color_discrete(name = "Division") +
  labs(
    x = "Time (Seconds)",
    y = "Workout Place",
    title = "Time to Complete 24.1 by Workout Place",
    subtitle = "Top 100 Overall Men and Women"
  )
```

I think this is a pretty cool plot in that it shows the "shape" of the performances by these athletes. The waterfall shape suggests that there's more separation (horizontal space) amongst the very top finishers, but the increasingly steep slope toward the right end of the graph suggests there's less time between worse-finishing places. Which makes sense.

See that one pink dot all the way to the left -- who crushed this workout that hard?

```{r}
wk1 |>
  slice_min(time_score, n = 1) |> 
  pull(athlete)
```


## 24.2

We'll do the same thing for 24.2. You can see the workout description [here](https://games.crossfit.com/workouts/open/2024/2)

```{r}
wk2 <- all_workouts |>
  filter(workout_num == 2) |>
  mutate(n_reps = parse_number(score))

wk2 |>
  ggplot(aes(x = n_reps, y = workout_place, color = div)) +
  geom_point(alpha = .5) +
  scale_y_reverse() +
  scale_color_discrete(name = "Division") +
  labs(
    x = "Reps",
    y = "Workout Place",
    title = "Reps Completed in 24.2 by Workout Place",
    subtitle = "Top 100 Overall Men and Women"
  )
 
```

We see basically the inverse pattern here -- the curve flattens toward the right (since more reps is better in this case). What's interesting is that we see this flattening occur at around the top ~100 finishers of the workout for women, but at maybe the top ~30 for men. This might have something to do with the scoring (1 double under obviously isn't equivalent to 1 deadlift), or it might be an indication that the top 100 women were considerably better than the next others in this workout. Again, it's kinda hard to tell.

## 24.3

And we'll wrap up by looking at the same type of plot for 24.3. You can see the workout description [here](https://games.crossfit.com/workouts/open/2024/3)

```{r}
wk3 <- all_workouts |>
  filter(workout_num == 3) |>
  mutate(time_score = period_to_seconds(ms(score)))

wk3 |>
  ggplot(aes(x = time_score, y = workout_place, color = div)) +
  geom_point(alpha = .5) +
  scale_y_reverse() +
  scale_color_discrete(name = "Division") +
  labs(
    x = "Time (Seconds)",
    y = "Workout Place",
    title = "Time to Complete 24.3 by Workout Place",
    subtitle = "Top 100 Overall Men and Women"
  )
```

Again we see basically the same pattern we saw in 24.1

I'll probably do something like this again for the Quarterfinals and Semis, so if there's anything else people would like to see, let me know.